---
title: "Count called variants in Desjardins Ashton dataset"
author: "Claudia Zirión-Martínez"
date: "2025-02-14"
format: 
    PrettyPDF-pdf:
        keep-tex: false
df-print: kable
editor: visual
embed-resources: true
code-overflow: wrap
---

## R Setup
```{r, output=FALSE}
library(tidyverse)
library(patchwork)
setwd("/FastData/czirion/Crypto_Diversity_Pipeline/analyses/ploidy/scripts")
```

## Desjardins

### Number of raw and filtered variants
Using bash count the number of variants called by Freebayes and filter by Snippy.  
Snippy filter are:   
**GT == 1/1**  
**QUAL >= 100**   
**DP >= 10**  
**A0/DP >= 0**  
```{bash, eval=FALSE}
cd /FastData/czirion/Crypto_Diversity_Pipeline/
tail -n +2 Crypto_Desjardins/config/metadata.csv | cut -d',' -f2 | while read line
do 
    raw=$(grep -v "#" Crypto_Desjardins/results/01.Samples/snippy/$line/snps.raw.vcf | wc -l)
    filt=$(grep -v "#" Crypto_Desjardins/results/01.Samples/snippy/$line/snps.filt.vcf | wc -l)
    echo $line,$raw,$filt >> analyses/ploidy/data/processed/snp_counts_desjardins.csv
done 
```

### Metadata
```{r, output=FALSE}
metadata <- read_csv("../../../Crypto_Desjardins/config/metadata.csv")
metadata <- metadata %>%
    select(sample, strain, lineage)
```
### Variant counts
```{r, message=FALSE}
snp_counts <- read_csv("../data/processed/snp_counts_desjardins.csv", col_names = c("sample", "n_raw", "n_filt"))

snp_counts <- snp_counts %>%
        mutate(n_removed = n_raw - n_filt, percent_filt = (n_filt / n_raw)*100)%>%
        left_join(metadata, by = "sample")
           
```
```{r, echo=FALSE}
#| fig-height: 7
#| fig-width: 7
#| fig-align: center
r <- ggplot(snp_counts, aes(x = n_raw, fill = lineage, color = lineage)) +
    geom_histogram(bins = 387) +
    theme_minimal() +
    labs(title = "Histograms of Raw and Filtered SNP Counts by Lineage", x = "Number of Raw SNPs", y = "Count", fill = "Lineage", color = "Lineage")
f <- ggplot(snp_counts, aes(x = n_filt, fill = lineage, color = lineage)) +
    geom_histogram(bins = 387) +
    theme_minimal() +
    labs(x = "Number of Filtered SNPs", y = "Count", fill = "Lineage", color = "Lineage") +
    theme(legend.position = "none")
r/f
```

```{r}
summary <- snp_counts %>%
    summarise(max_raw = max(n_raw),
              max_filtered = max(n_filt),
              max_removed = max(n_removed),
              median_raw = median(n_raw),
              median_filt = median(n_filt),
              median_removed = median(n_removed))
summary   
```
```{r}
sorted <- snp_counts %>%
    arrange(desc(n_removed))
head(sorted, 30)
```


## Ashton
### Number of raw and filtered variants
Using bash count the number of variants called by Freebayes and filter by Snippy.  
Snippy filter are:   
**GT == 1/1**  
**QUAL >= 100**   
**DP >= 10**  
**A0/DP >= 0**  
```{bash, eval=FALSE}
cd /FastData/czirion/Crypto_Diversity_Pipeline/
tail -n +2 Crypto_Ashton/config/metadata.csv | cut -d',' -f2 | while read line
do 
    raw=$(grep -v "#" Crypto_Ashton/results/01.Samples/snippy/$line/snps.raw.vcf | wc -l)
    filt=$(grep -v "#" Crypto_Ashton/results/01.Samples/snippy/$line/snps.filt.vcf | wc -l)
    echo $line,$raw,$filt >> analyses/ploidy/data/processed/snp_counts_ashton.csv
done 
```
### Metadata
```{r, output=FALSE}
metadata <- read_csv("../../../Crypto_Ashton/config/metadata.csv")
metadata <- metadata %>%
    select(sample, strain, lineage, VNI_subdivision)
```
### Variant counts
```{r, output=FALSE}
snp_counts <- read_csv("../data/processed/snp_counts_ashton.csv", col_names = c("sample", "n_raw", "n_filt"))
```
```{r}

snp_counts <- snp_counts %>%
        mutate(n_removed = n_raw - n_filt, percent_filt = (n_filt / n_raw)*100)%>%
        left_join(metadata, by = "sample")
```

```{r, echo=FALSE}
#| fig-height: 7
#| fig-width: 7
#| fig-align: center
r <- ggplot(snp_counts, aes(x = n_raw, fill = VNI_subdivision, color = VNI_subdivision)) +
    geom_histogram(bins = 387) +
    theme_minimal() +
    labs(title = "Histograms of Raw and Filtered SNP Counts by Sublineage", x = "Number of Raw SNPs", y = "Count", fill = "Sublineage", color = "Sublineage")
f <- ggplot(snp_counts, aes(x = n_filt, fill = VNI_subdivision, color = VNI_subdivision)) +
    geom_histogram(bins = 387) +
    theme_minimal() +
    labs(x = "Number of Filtered SNPs", y = "Count", fill = "Sublineage", color = "Sublineage") +
    theme(legend.position = "none")
r/f
```

```{r}
summary <- snp_counts %>%
    summarise(max_raw = max(n_raw),
              max_filtered = max(n_filt),
              max_removed = max(n_removed),
              median_raw = median(n_raw),
              median_filt = median(n_filt),
              median_removed = median(n_removed))
summary
```
```{r}
sorted <- snp_counts %>%
    arrange(desc(n_removed))
head(sorted, 30)
```