# Comparison of WeavePop CNVs with PerSVade and CNVpytor

## Setup {#setup}

```{r, eval=FALSE, echo=FALSE}
setwd("/FastData/czirion/WeavePop_Cneoformans/analyses")
```

Libraries

```{r libraries, output=FALSE}
library(tidyverse)
library(ggbeeswarm)
library(patchwork)
library(ggnewscale)
library(RColorBrewer)
library(svglite)
library(scales)
```
Paths
```{r, warning = FALSE, message = FALSE}
samples_path = "results/tables/haploid_samples_with_cnv.tsv"
name_c = "cnvpytor"
name_p = "HMMcopy_AneuFinder"
```

Read table of metrics and CNV calls per window of all samples and all softwares and create
one dataframe with the agreement of all pairs of softwares.

`w` = WeavePop
`c` = CNVpytor
`p` = PerSVade

The `type` columns have the call of the software in the left of the name to the left of `_2_` and the call of the software in the right of the name to the rigth of `_2_`.

```{r, warning = FALSE, message = FALSE}
s <- read_delim(samples_path, delim= '\t', col_names = c("sample", "dataset", "lineage"))
samples = pull(s, sample)
types = c("single_copy_2_single_copy", "duplication_2_duplication", "deletion_2_deletion",
        "single_copy_2_duplication" ,"duplication_2_single_copy" ,
        "single_copy_2_deletion", "deletion_2_single_copy",
        "deletion_2_duplication","duplication_2_deletion")

df_all <- data.frame()
for (sample in samples){
    weavepop_path = paste("data/cnv_software/depth_by_windows/", sample ,".tsv", sep = "")
    c_path = paste("results/cnv_software/results_", name_c, "/", sample ,"_500_windows.bed", sep = "")
    p_path = paste("results/cnv_software/results_", name_p, "/", sample ,"_500_windows.bed", sep = "")

    weavepop <- read_delim(weavepop_path, delim= '\t', col_names = TRUE)%>%
        mutate(start_1 = start + 1)%>%
        select(accession, start = start_1, end, smooth_depth, cnv, repeat_overlap_bp)

    c <- read_delim(c_path, delim= '\t', col_names = c("accession", "start", "end", "norm_depth_c", "cnv_c"))%>%
        mutate(cnv_c = ifelse(cnv_c == ".", "single_copy", cnv_c),
            norm_depth_c = as.numeric(ifelse(norm_depth_c == ".", "1", norm_depth_c)))

    p <- read_delim(p_path, delim= '\t', col_names = c("accession", "start", "end", "norm_depth_p", "cnv_p"))%>%
        mutate(cnv_p = ifelse(cnv_p == ".", "single_copy", cnv_p),
            norm_depth_p = as.numeric(ifelse(norm_depth_p == ".", "1", norm_depth_p)))

    df <- left_join(weavepop, c, by = c("accession", "start", "end"))%>%
        left_join(., p, by = c("accession", "start", "end"))%>%
        mutate(sample = sample,
                agreement_wc = ifelse(cnv == cnv_c, "agree", "disagree"),
                agreement_wp = ifelse(cnv == cnv_p, "agree", "disagree"),
                agreement_cp = ifelse(cnv_c == cnv_p, "agree", "disagree"),
                type_wc = paste(cnv, cnv_c, sep = "_2_"),
                type_wp = paste(cnv, cnv_p, sep = "_2_"),
                type_cp = paste(cnv_c, cnv_p, sep = "_2_"),
                total = n())%>%
        mutate(across(where(is.character), as.factor))
    
    df$type_wc <- factor(df$type_wc, levels = types)
    df$type_wp <- factor(df$type_wp, levels = types)
    df$type_cp <- factor(df$type_cp, levels = types)

    df_all <- bind_rows(df_all, df)
}
```

## WeavePop vs PerSVade

```{r, warning = FALSE, message = FALSE}
s_wp <-df_all %>%
        group_by(sample, agreement_wp, cnv, cnv_p, type_wp, total)%>%
        summarize(
            repeat_fraction = mean(repeat_overlap_bp / 500),
                    n = n())%>%
        mutate(n_pct = n / total * 100)%>%
        select(-total)
```
```{r, warning = FALSE, message = FALSE}
s_wp %>%
    group_by(agreement_wp, type_wp)%>%
    summarize(mean_repeat_fraction = mean(repeat_fraction),
                mean_n = mean(n),
                mean_n_pct = mean(n_pct))
```
```{r, warning = FALSE, message = FALSE}
a_wp <- df_all %>%
    group_by(sample,agreement_wp, total)%>%
    summarize(repeat_fraction = mean(repeat_overlap_bp / 500),
              n = n())%>%
        mutate(n_pct = n / total * 100)
```

```{r, warning = FALSE, message = FALSE}
a_wp %>%
    group_by(agreement_wp)%>%
    summarize(mean_repeat_fraction = mean(repeat_fraction),
                mean_n = mean(n),
                mean_n_pct = mean(n_pct))
```
```{r, warning = FALSE, message = FALSE}
ggplot(s_wp)+
    geom_quasirandom(aes(x = type_wp, y = repeat_fraction))+
    geom_boxplot(aes(x = type_wp, y = repeat_fraction), alpha = 0.2)+
    facet_grid(~agreement_wp, scale = "free")+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    labs(title = "WeavePop vs PerSVade",
        x = "",
        y = "Repeat fraction")
```
```{r, warning = FALSE, message = FALSE}
ggplot(filter(s_wp, type_wp != "single_copy_2_single_copy"))+
    geom_quasirandom(aes(x = type_wp, y = n))+
    geom_boxplot(aes(x = type_wp, y = n), alpha = 0.2)+
    facet_grid(~agreement_wp, scale = "free")+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    labs(title = "WeavePop vs PerSVade",
        x = "",
        y = "Number of windows")
```

## WeavePop vs CNVpytor
```{r, warning = FALSE, message = FALSE}
s_wc <-df_all %>%
        group_by(sample, agreement_wc, cnv, cnv_c, type_wc, total)%>%
        summarize(
            repeat_fraction = mean(repeat_overlap_bp / 500),
                    n = n())%>%
        mutate(n_pct = n / total * 100)%>%
        select(-total)
```
```{r, warning = FALSE, message = FALSE}
s_wc %>%
    group_by(agreement_wc, type_wc)%>%
    summarize(mean_repeat_fraction = mean(repeat_fraction),
                mean_n = mean(n),
                mean_n_pct = mean(n_pct))
```
```{r, warning = FALSE, message = FALSE}
a_wc <- df_all %>%
    group_by(sample,agreement_wc, total)%>%
    summarize(repeat_fraction = mean(repeat_overlap_bp / 500),
              n = n())%>%
        mutate(n_pct = n / total * 100)
```

```{r, warning = FALSE, message = FALSE}
a_wc %>%
    group_by(agreement_wc)%>%
    summarize(mean_repeat_fraction = mean(repeat_fraction),
                mean_n = mean(n),
                mean_n_pct = mean(n_pct))
```
```{r, warning = FALSE, message = FALSE}
ggplot(s_wc)+
    geom_quasirandom(aes(x = type_wc, y = repeat_fraction))+
    geom_boxplot(aes(x = type_wc, y = repeat_fraction), alpha = 0.2)+
    facet_grid(~agreement_wc, scale = "free")+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    labs(title = "WeavePop vs CNVPytor",
        x = "",
        y = "Repeat fraction")
```
```{r, warning = FALSE, message = FALSE}
ggplot(filter(s_wc, type_wc != "single_copy_2_single_copy"))+
    geom_quasirandom(aes(x = type_wc, y = n))+
    geom_boxplot(aes(x = type_wc, y = n), alpha = 0.2)+
    facet_grid(~agreement_wc, scale = "free")+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    labs(title = "WeavePop vs CNVpytor",
        x = "",
        y = "Number of windows")
```

## CNVpytor vs PerSVade
```{r, warning = FALSE, message = FALSE}
s_cp <-df_all %>%
        group_by(sample, agreement_cp, cnv_c, cnv_p, type_cp, total)%>%
        summarize(
            repeat_fraction = mean(repeat_overlap_bp / 500),
                    n = n())%>%
        mutate(n_pct = n / total * 100)%>%
        select(-total)
```
```{r, warning = FALSE, message = FALSE}
s_cp %>%
    group_by(agreement_cp, type_cp)%>%
    summarize(mean_repeat_fraction = mean(repeat_fraction),
                mean_n = mean(n),
                mean_n_pct = mean(n_pct))
```
```{r, warning = FALSE, message = FALSE}
a_cp <- df_all %>%
    group_by(sample,agreement_cp, total)%>%
    summarize(repeat_fraction = mean(repeat_overlap_bp / 500),
              n = n())%>%
        mutate(n_pct = n / total * 100)
```

```{r, warning = FALSE, message = FALSE}
a_cp %>%
    group_by(agreement_cp)%>%
    summarize(mean_repeat_fraction = mean(repeat_fraction),
                mean_n = mean(n),
                mean_n_pct = mean(n_pct))
```
```{r, warning = FALSE, message = FALSE}
ggplot(s_cp)+
    geom_quasirandom(aes(x = type_cp, y = repeat_fraction))+
    geom_boxplot(aes(x = type_cp, y = repeat_fraction), alpha = 0.2)+
    facet_grid(~agreement_cp, scale = "free")+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    labs(title = "CNVpytor vs PerSVade",
        x = "",
        y = "Repeat fraction")
```
```{r, warning = FALSE, message = FALSE}
ggplot(filter(s_cp, type_cp != "single_copy_2_single_copy"))+
    geom_quasirandom(aes(x = type_cp, y = n))+
    geom_boxplot(aes(x = type_cp, y = n), alpha = 0.2)+
    facet_grid(~agreement_cp, scale = "free")+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    labs(title = "CNVpytor vs PerSVade",
        x = "",
        y = "Number of windows")
```
