---
title: "Detect chromosomal duplications"
author: "Claudia Zirión-Martínez"
date: "2025-02-13"
format: 
    PrettyPDF-pdf:
        keep-tex: false
df-print: kable
editor: visual
embed-resources: true
code-overflow: wrap
---
## Setup
```{r, output=FALSE}
library(tidyverse)
library(ggtree)
library(ggtreeExtra)
library(ape)
library(ggnewscale)
library(RColorBrewer)
setwd("/FastData/czirion/Crypto_Diversity_Pipeline/analyses/tree_duplications/scripts")
```

## Metadata
Load the necessary data
```{r}
metadata <- read.csv(
    "../../data/processed/metadata_ashton_desj_all_fungalpop_H99.csv",
    header = TRUE)
```

Get one dataframe for each variable to be plotted as a separate metadata column in the tree
```{r}
lineage <- metadata %>%
    select(strain, lineage)%>%
    column_to_rownames("strain")

source <- metadata %>%
    select(strain, source)%>%
    column_to_rownames("strain")

sublineage <- metadata %>%
    select(strain, vni_subdivision)%>%
    column_to_rownames("strain")

dataset <- metadata %>%
    select(strain, dataset)%>%
    column_to_rownames("strain")
```

## Duplications
```{r}
duplications <- read.delim(
    "../results/tables/duplications_polished.tsv",
    sep = "\t", header = TRUE, stringsAsFactors = TRUE)
```

```{r}
duplications_full <- duplications %>%
    select(strain, chromosome) %>%
    distinct()
```

Make matrix of duplicated chromosomes
```{r}
dup_chroms <- duplications_full %>%
    select(strain, chromosome)%>%
    mutate(duplicated_full = 1)%>%
    arrange(chromosome)%>%
    pivot_wider(names_from = chromosome, values_from = duplicated_full, values_fill = 0)%>%
    column_to_rownames("strain")%>%
    mutate(across(everything(), ~ ifelse(. == 1, cur_column(),"Euploid")))

euploid_strain <- metadata %>%
    filter(!strain %in% duplications_full$strain)%>%
    select(strain)

for (chrom in colnames(dup_chroms)){
    euploid_strain[chrom] <- "Euploid"
}

dup_chroms <- euploid_strain %>%
    column_to_rownames("strain") %>%
    bind_rows(dup_chroms)
```


## Tree
```{r}
merged_tree_path <- "/FastData/czirion/Crypto_Diversity_Pipeline/analyses/data/processed/merged_tree.newick"
tree <- read.tree(merged_tree_path)
```
Remove tips that are not in metadata$strain

```{r}
tree <- drop.tip(tree, setdiff(tree$tip.label, metadata$strain))
```

## Plots
```{r, eval=FALSE, echo=FALSE}
display colorblind friendly palettes
display.brewer.all(colorblindFriendly = TRUE)
display.brewer.all(colorblindFriendly = FALSE)
```

Create vectors of colors for each metadata variable with the name of the levels as names of the colors
```{r}
dataset_colors <- c(brewer.pal(9, "Set1")[c(1, 2)], "white")
names(dataset_colors) <- levels(as.factor(dataset$dataset))

lineage_colors <- brewer.pal(8, "Dark2")[c(1, 2, 3, 4)]
names(lineage_colors) <- levels(as.factor(lineage$lineage))

sublineage_colors <- c(brewer.pal(12, "Set3")[c(1:9)])
names(sublineage_colors) <- levels(sublineage$vni_subdivision)

source_colors <- brewer.pal(11, "BrBG")[c(9, 3)] # 9, 3 are the colors for the two sources
names(source_colors) <- levels(as.factor(source$source))

chrom_colors <- c(brewer.pal(nlevels(duplications$chromosome), "Paired"), "grey93")
names(chrom_colors) <- c(levels(duplications$chromosome), "Euploid")
```

### Tree of all samples with duplications of all chromosomes

```{r, message=FALSE, warning=FALSE, echo=FALSE}
#| fig-height: 7
#| fig-width: 7
#| fig-align: center
m <- ggtree(tree, layout = "circular", size = 0.1, branch.length = "none") +  
        geom_tiplab(aes(label = label), size = 0.5, align =TRUE, 
                    linetype = "dashed", linesize = .03)

m1 <- gheatmap(m, dataset, width=.05, colnames=FALSE, offset=2) +
            scale_fill_manual(values = dataset_colors, name="Dataset", na.translate = FALSE)+
            guides(fill = guide_legend(order = 1))+
            new_scale_fill()

m2 <- gheatmap(m1, lineage, width=.05, colnames=FALSE, offset=4) +
            scale_fill_manual(values = lineage_colors, name="Lineage", na.translate = FALSE)+
            guides(fill = guide_legend(order = 2))+
            new_scale_fill()

m3 <- gheatmap(m2, sublineage, width=.05, colnames=FALSE, offset=6) +
            scale_fill_manual(values = sublineage_colors,
                name="VNI Sublineage",
                na.translate = FALSE,
                limits = levels(sublineage$vni_subdivision))+
            guides(fill = guide_legend(order = 3))+
            new_scale_fill()

m4 <- gheatmap(m3, source, width=.05, colnames=FALSE, offset=8) +
        scale_fill_manual(values = source_colors, name="Source", na.translate = FALSE)+
        guides(fill = guide_legend(order = 4))+
        new_scale_fill()

m5 <- gheatmap(m4, dup_chroms, width=.32, colnames = FALSE, offset=11,) +
    scale_fill_manual(values = chrom_colors, name="Duplicated\nchromosomes", na.translate = FALSE )+
    guides(fill = guide_legend(order = 5))+
    theme(legend.position = "right",
        legend.direction = "vertical",
        legend.title = element_text(size=9),
        legend.text=element_text(size=7),
        legend.key.size = unit(0.3, "cm"),
        plot.margin = margin(0, 0, 0, 0, "cm"))
m5
```

```{r, eval=FALSE, echo=FALSE}
ggsave("/FastData/czirion/Crypto_Diversity_Pipeline/analyses/tree_duplications/results/figures/tree_merged_duplications.png", p5, height = 7, width = 7, units = "in", dpi = 900)
```

### Tree of all samples with duplications of chromosomes 12 and 13

Subset the duplications_full data frame to only include strains with duplications of chromosomes 12 and 13

```{r}
dup_chroms_12_13 <- dup_chroms %>%
    select(chr12, chr13)
```
```{r, message=FALSE, warning=FALSE, echo=FALSE}
#| fig-height: 5
#| fig-width: 7
#| fig-align: center
p5 <- gheatmap(m4, dup_chroms_12_13, width=.1, colnames = FALSE, offset=11,) +
    scale_fill_manual(values = chrom_colors, name="Duplicated\nchromosomes", na.translate = FALSE )+
    guides(fill = guide_legend(order = 5))+
    theme(legend.position = "right",
        legend.direction = "vertical",
        legend.title = element_text(size=9),
        legend.text=element_text(size=7),
        legend.key.size = unit(0.3, "cm"),
        plot.margin = margin(0, 0, 0, 0, "cm"))
p5
```

```{r, echo=FALSE, eval=FALSE}
ggsave("/FastData/czirion/Crypto_Diversity_Pipeline/analyses/tree_duplications/results/figures/tree_merged_duplications_12_13.png", p5, height = 7, width = 7, units = "in", dpi = 900)
```

### Tree with only the samples that have duplications and the references

```{r}
keep_strains <- c(levels(duplications_full$strain), "H99", "Bt22", "Bt81")
tree_dups <- drop.tip(tree, setdiff(tree$tip.label, keep_strains))
```

#### Dataset, lineage, sublineage, source, duplications
```{r, message=FALSE, warning=FALSE, echo=FALSE}
#| fig-height: 7
#| fig-width: 9
#| fig-align: center
p <- ggtree(tree_dups, layout = "rectangular", size = 0.5, branch.length = "none") + 
    geom_tiplab(aes(label = label), size = 3, align =TRUE, 
                    linetype = "dashed", linesize = 0.1, offset = 1)

p1 <- gheatmap(p, dataset, width=0.1, colnames=FALSE, offset=8) +
    scale_fill_manual(values = dataset_colors, name="Dataset", na.translate = FALSE)+
    guides(fill = guide_legend(order = 1))+
    new_scale_fill()

p2 <- gheatmap(p1, lineage, width=0.1, colnames=FALSE, offset=9.5) +
    scale_fill_manual(values = lineage_colors, name="Lineage", na.translate = FALSE)+
    guides(fill = guide_legend(order = 2))+
    new_scale_fill()

p3 <- gheatmap(p2, sublineage, width=0.1, colnames=FALSE, offset=11) +
    scale_fill_manual(values = sublineage_colors, name="VNI Sublineage", na.translate = FALSE)+ 
    guides(fill = guide_legend(order = 3))+
    new_scale_fill()

p4 <- gheatmap(p3, source, width=0.1, colnames=FALSE, offset=12.5) +
        scale_fill_manual(values = source_colors, name="Source", na.translate = FALSE)+
        guides(fill = guide_legend(order = 4))+
        new_scale_fill()

p5 <- gheatmap(p4, dup_chroms, width=0.7, colnames = FALSE, offset=14) +
    scale_fill_manual(values = chrom_colors, name="Duplicated\nchromosomes", na.translate = FALSE )+
    guides(fill = guide_legend(order = 5))+
    theme(legend.position = "right",
        legend.direction = "vertical",
        legend.title = element_text(size=9),
        legend.text=element_text(size=7),
        legend.key.size = unit(0.3, "cm"))
p5
```

```{r, echo=FALSE, eval=FALSE}
ggsave("/FastData/czirion/Crypto_Diversity_Pipeline/analyses/tree_duplications/results/figures/tree_merged_duplications_only_duplicated.png", p5, height = 7, width = 9, units = "in", dpi = 900)
```

#### Lineage, sublineage, duplications

```{r, message=FALSE, warning=FALSE, echo=FALSE}
#| fig-height: 7
#| fig-width: 9
#| fig-align: center
p <- ggtree(tree_dups, layout = "rectangular", size = 0.5, branch.length = "none") + 
    geom_tiplab(aes(label = label), size = 3, align =TRUE, 
                    linetype = "dashed", linesize = 0.1, offset = 1)
p1 <- gheatmap(p, lineage, width=0.1, colnames=FALSE, offset=8) +
    scale_fill_manual(values = lineage_colors, name="Lineage", na.translate = FALSE)+
    guides(fill = guide_legend(order = 2))+
    new_scale_fill()

p2 <- gheatmap(p1, sublineage, width=0.1, colnames=FALSE, offset=10) +
    scale_fill_manual(values = sublineage_colors, name="VNI Sublineage", na.translate = FALSE)+ 
    guides(fill = guide_legend(order = 3))+
    new_scale_fill()

p3 <- gheatmap(p2, dup_chroms, width=0.7, colnames = FALSE, offset=12) +
    scale_fill_manual(values = chrom_colors, name="Duplicated\nchromosomes", na.translate = FALSE )+
    guides(fill = guide_legend(order = 5))+
    theme(legend.position = "right",
        legend.direction = "vertical",
        legend.title = element_text(size=9),
        legend.text=element_text(size=7),
        legend.key.size = unit(0.3, "cm"))
p3
```

```{r, echo=FALSE, eval=FALSE}
ggsave("/FastData/czirion/Crypto_Diversity_Pipeline/analyses/tree_duplications/results/figures/tree_merged_duplications_only_duplicated2.png", p3, height = 7, width = 9, units = "in", dpi = 900)
```

#### Lineage, duplications, sublineage

```{r, message=FALSE, warning=FALSE, echo=FALSE}
#| fig-height: 7
#| fig-width: 9
#| fig-align: center
p <- ggtree(tree_dups, layout = "rectangular", size = 0.5, branch.length = "none") + 
    geom_tiplab(aes(label = label), size = 3, align =TRUE, 
                    linetype = "dashed", linesize = 0.1, offset = 1)

p1 <- gheatmap(p, lineage, width=0.1, colnames=FALSE, offset=8) +
    scale_fill_manual(values = lineage_colors, name="Lineage", na.translate = FALSE)+
    guides(fill = guide_legend(order = 2))+
    new_scale_fill()

p2 <- gheatmap(p1, dup_chroms, width=0.7, colnames = FALSE, offset=10) +
    scale_fill_manual(values = chrom_colors, name="Duplicated\nchromosomes", na.translate = FALSE )+
    guides(fill = guide_legend(order = 2))+
    new_scale_fill()

p3 <- gheatmap(p2, sublineage, width=0.1, colnames=FALSE, offset=19.5) +
    scale_fill_manual(values = sublineage_colors, name="VNI Sublineage", na.translate = FALSE)+ 
    guides(fill = guide_legend(order = 3))+
    theme(legend.position = "right",
        legend.direction = "vertical",
        legend.title = element_text(size=9),
        legend.text=element_text(size=7),
        legend.key.size = unit(0.3, "cm"))
p3
```

```{r, echo=FALSE, eval=FALSE}
ggsave("/FastData/czirion/Crypto_Diversity_Pipeline/analyses/tree_duplications/results/figures/tree_merged_duplications_only_duplicated3.png", p3, height = 7, width = 9, units = "in", dpi = 900)
```