---
title: "Detect chromosomal duplications"
author: "Claudia Zirión-Martínez"
date: "2025-02-13"
format: 
    PrettyPDF-pdf:
        keep-tex: false
df-print: kable
editor: visual
embed-resources: true
code-overflow: wrap
---

## Setup
```{r, message=FALSE}
setwd("/FastData/czirion/Crypto_Diversity_Pipeline/analyses/tree_duplications/scripts")
library(tidyverse)
```

## Detect partial duplications
After manual inspection of plots (see scripts `duplications_detect.qmd` `duplications_gather_plots.xsh`) I found that the following samples have partial duplications instead of full chromosome duplications.

```{r}
partial <- data.frame(sample= 
                            c("ERS2541051",
                            "ERS2541051",
                            "ERS542397",
                            "ERS1142798",
                            "ERS542490",
                            "ERS1142878",
                            "ERS2541358",
                            "SRS881238",
                            "ERS542397",
                            "ERS542498"),
                    chromosome = 
                            c("chr09",
                            "chr12",
                            "chr02",
                            "chr14",
                            "chr04",
                            "chr09",
                            "chr02",
                            "chr09",
                            "chr04",
                            "chr04"))

partial$sample_chromosome <- paste(partial$sample, partial$chromosome, sep="_")
```

## Remove partial duplications

```{r, message=FALSE}
duplications_putative <- read_tsv("../results/tables/duplications_putative.tsv")

duplications <- duplications_putative %>%
    mutate(sample_chromosome = paste(sample, chromosome, sep="_"))%>%
    filter(!sample_chromosome %in% partial$sample_chromosome)

duplicated_full_strain <- duplications %>%
    select(dataset, lineage, sample, strain,
            source, chromosome, 
            norm_chrom_median, percent_cnv_size)
```

```{r}
write_tsv(duplicated_full_strain,
    "../results/tables/duplications_polished.tsv")
```
## Get multiple summary tables

Number of duplicated chromosomes per sample.
```{r,message=FALSE}
dup_sample <- duplications %>%
    group_by(dataset,lineage, sample, strain, source) %>%
    summarise(n_chroms = n_distinct(chromosome), chromosomes = paste(chromosome, collapse = ", ")) %>%
    arrange(desc(n_chroms))
dup_sample
```

Number of samples with duplications in each group of dataset-lineage-chromosome.
```{r, message=FALSE}
dup_dataset_lineage_chromosome <- duplications %>%
    group_by(dataset,lineage, chromosome) %>%
    summarise(n_samples = n_distinct(sample), samples_in_dataset_lineage = first(samples_in_dataset_lineage))%>%
    mutate(percent_samples = round((n_samples / samples_in_dataset_lineage) * 100, 1))%>%
    select(dataset,lineage, chromosome, n_samples, samples_in_dataset_lineage, percent_samples)%>%
    arrange(chromosome, desc(lineage), desc(n_samples))
dup_dataset_lineage_chromosome
```

Number of samples with duplications in each group of lineage-chromosome.
```{r,message=FALSE}
dup_lineage_chromosome <- duplications%>%
    group_by(lineage, chromosome) %>%
    summarise(n_samples = n_distinct(sample), samples_in_lineage = first(samples_in_lineage))%>%
    mutate(percent_samples = round((n_samples / samples_in_lineage) * 100, 1))%>%
    select(lineage, chromosome, n_samples, samples_in_lineage,percent_samples)%>%
    arrange(chromosome, desc(lineage), desc(n_samples))
dup_lineage_chromosome 
```

Number of samples with duplications in each group of lineage-datset.
```{r, message=FALSE}
dup_lineage_dataset <- duplications%>%
    group_by(dataset,lineage) %>%
    summarise(n_samples = n_distinct(sample), samples_in_dataset_lineage = first(samples_in_dataset_lineage))%>%
    mutate(percent_samples = round((n_samples / samples_in_dataset_lineage) * 100, 1))%>%
    select(lineage, n_samples, samples_in_dataset_lineage, percent_samples)%>%
    arrange(desc(lineage), desc(n_samples))
dup_lineage_dataset

```

Number of samples with duplications in each chromosome.
```{r, message=FALSE}
dup_chromosome <- duplications %>%
    group_by(chromosome) %>%
    summarise(n_samples = n_distinct(sample), total_samples = first(total_samples))%>%
    mutate(percent_samples = round((n_samples / total_samples) * 100, 1))%>%
    select(chromosome, n_samples,total_samples, percent_samples)%>%
    arrange(chromosome, desc(n_samples))
dup_chromosome 
```