---
title: "CNV and Depth Metrics of Chromosomes"
author: "Claudia Zirión-Martínez"
date: "2025-05-01"

fig-height: 7
fig-width: 9

format: 
  html:
    page-layout: full
---


## Setup {#setup}

```{r, eval=FALSE, echo=FALSE}
setwd("/FastData/czirion/WeavePop_Cneoformans/analyses")
```

Libraries

```{r libraries, output=FALSE}
library(tidyverse)
library(patchwork)
library(ggbeeswarm)
library(ggnewscale)
library(RColorBrewer)
library(hexbin)
library(ggExtra)
library(ggrepel)
library(truncnorm)
library(ggpattern)
library(svglite)
```

Paths

```{r paths}
metadata_path <-
    "data/processed/metadata_ashton_desj_all_weavepop_final_H99.csv"
cnv_chroms_path <-
    "../Crypto_Desjardins_Ashton/results_filtered/02.Dataset/cnv/cnv_chromosomes.tsv"
chrom_metrics_out_path <-
    "results/tables/chrom_metrics.tsv"
ploidy_path <- "results/tables/ploidy_from_plots.tsv"

```

## Prepare dataset {#prepare-dataset}

### Metadata {#metadata}

Use the metadata table that has all the samples included in the final Crypto_Desjardins_Ashton dataset (n = 1055) and H99.  
Select needed columns, remove H99 and get the number of samples per dataset and lineage, per lineage, and total.

```{r, message=FALSE, warning=FALSE}
metadata <- read.delim(
    metadata_path,
    header=TRUE,
    sep=",",
    stringsAsFactor = TRUE)
metadata <- metadata %>%
    select(sample, strain, source, lineage, dataset, vni_subdivision)%>%
    filter(!strain == "H99") %>%
    group_by(dataset, lineage)%>%
    mutate(samples_in_dataset_lineage = n_distinct(sample))%>%
    ungroup() %>%
    group_by(lineage)%>%
    mutate(samples_in_lineage = n_distinct(sample))%>%
    ungroup()%>%
    mutate(total_samples = n_distinct(sample))%>%
    droplevels()
```

```{r}
ploidy <- read.delim(
                ploidy_path,
                header = TRUE,
                sep = "\t",
                stringsAsFactors = FALSE)%>%
          select(sample, lineage,
                 accession, chromosome, ploidy, type, gappy, 
                 distribution_pattern, fraction_duplications,
                 steps, smile, uniform)%>%
            distinct()
    
ploidy$ploidy <- factor(ploidy$ploidy, levels = c("haploid", "diploid", "triploid", "tetraploid"))
ploidy$type <- factor(ploidy$type, levels = c("full", "large", "partial"))
```

## CNV Regions and Depth per Chromosome

```{r}
cnv_chroms <- read.delim(
    cnv_chroms_path,
    header = TRUE,
    sep = "\t")%>%
    mutate(chromosome = str_pad(chromosome, 2, pad = "0"))
cnv_chroms$chromosome <- as.factor(cnv_chroms$chromosome)

levels(cnv_chroms$chromosome) <- paste("chr", levels(cnv_chroms$chromosome), sep="")
```
Calculate   
* `percent_largest_region` : Percentage of length of chromosome covered by the largest CNV region of the chormosome.  
* `diff_span_cov`: Difference between `span_percent` and `coverage_percent`.  
* `n_regions_100kb`: Number of regions per 100kb  
```{r}
cnv_chroms <- cnv_chroms %>%
    mutate(percent_largest_region = (size_largest_region / length)*100,
            diff_span_cov = abs(span_percent - coverage_percent),
            n_regions_100kb = (n_regions / length) * 100000)
```
Pivot table to get only one line per chromosome.  
Join with depth per chromosome and with ploidy informations. 
```{r}
chrom_metrics <- cnv_chroms %>%
    pivot_wider(names_from = cnv,
                values_from = c("n_regions","total_size_regions","coverage_percent",  
                            "span_percent","diff_span_cov","size_smallest_region","size_largest_region",
                            "percent_largest_region", "n_regions_100kb",
                            "std_regions_size","norm_depth_mean","norm_depth_median",   
                            "smooth_depth_mean","smooth_depth_median")) %>%
    left_join(ploidy, by = c("sample", "accession", "lineage", "chromosome"))%>%
    left_join(metadata, by = c("sample", "lineage"))
```

### Distribution

```{r}
ggplot(chrom_metrics)+
        geom_quasirandom(aes(x = chromosome, y = coverage_percent_duplication))
```
```{r}
ggplot(filter(chrom_metrics, coverage_percent_duplication > 0))+
        geom_quasirandom(aes(x = chromosome, y = coverage_percent_duplication), alpha = 0.5)+
        ylim(0,10)
```
```{r}
ggplot(filter(chrom_metrics, coverage_percent_duplication > 0), aes(x = lineage, y = coverage_percent_duplication))+
        geom_quasirandom()+
        geom_boxplot()
```
```{r}
ggplot(filter(chrom_metrics, coverage_percent_duplication > 0),
         aes(x = lineage, y = coverage_percent_duplication))+
        geom_quasirandom(alpha = 0.5)+
        ylim(0,10)
```
```{r}
ggplot(chrom_metrics)+
        geom_quasirandom(aes(x = chromosome, y = total_size_regions_duplication))
```
```{r}
ggplot(filter(chrom_metrics, total_size_regions_duplication > 0))+
        geom_quasirandom(aes(x = chromosome, y = total_size_regions_duplication), alpha = 0.5)+
        ylim(0,100000)
```
```{r}
ggplot(filter(chrom_metrics, total_size_regions_duplication > 0), aes(x = lineage, y = total_size_regions_duplication))+
        geom_quasirandom()+
        geom_boxplot()
```
```{r}
ggplot(filter(chrom_metrics, total_size_regions_duplication > 0),
         aes(x = lineage, y = total_size_regions_duplication))+
        geom_quasirandom(alpha = 0.5)+
        ylim(0,100000)
```


### Simulate expected trend

Simulation of haploid chromosomes with all proportions (in steps of 5) of
percentage of chromosome covered by deletions, duplications and single-copy regions.

Select simulation of percentage of deletion = 5.

```{r}
n <- 100000
sd_v <- 0.2

p <- data.frame(p_del = c(rep(seq(0,100, by = 5), each = 100)),
                p_sc  = c(rep(seq(0,100, by = 5), 100)))%>%
    mutate(sum = p_del + p_sc)%>%
    filter(sum <= 100)%>%
    distinct()
               
sample_summary <- p %>%
        mutate(p_dup = 100 - p_del - p_sc,
                n_del = p_del * n / 100,
                n_sc = p_sc * n  / 100,
                n_dup = p_dup * n  / 100)%>%
        rowwise()%>%
        mutate(median_depth = median(c(rtruncnorm(n = n_del, a = 0, mean = 0, sd = sd_v),
                                 rnorm(n_sc, mean = 1, sd = sd_v), 
                                 rnorm(n_dup, mean = 2, sd = sd_v))),
                mean_depth = mean(c(rtruncnorm(n = n_del, a = 0, mean = 0, sd = sd_v),
                                 rnorm(n_sc, mean = 1, sd = sd_v), 
                                 rnorm(n_dup, mean = 2, sd = sd_v))))%>%
        select(p_del,p_sc,p_dup, median_depth, mean_depth)%>%
        mutate(category = paste(p_del,p_sc,p_dup, sep = "_"),
                dup_filter = ifelse(p_dup >= 50, "half_duplication", "small_duplication"),
                del_filter = ifelse(p_del >= 50, "half_deletion", "small_deletion"))

trend <- filter(sample_summary, p_del == 5)
```

```{r, echo=FALSE, eval = FALSE}
        # geom_point(aes(shape = gappy, color = type)) +
        # scale_color_brewer(name = "Type", palette = "Set1") +

        # geom_abline(intercept = 1, slope = 0.01, color = "gray", linetype = "solid") +
        # geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")
        # geom_text_repel(aes(label = paste(sample, chromosome)), size = 3, max.overlaps = 7) +
```
### Base plot
```{r, message = FALSE, warning = FALSE}
ggplot() + theme_minimal()
p <- ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_point()+
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom", legend.direction = "vertical")+
        labs(y = "Normalized Chromosome Median",
            x = "Percent of Chromosome Covered by Duplications")
p1 <- ggMarginal(
    p,
    type = "boxplot",
    margins = "both",
    size = 5,
    groupColour = FALSE,
    groupFill = FALSE
)      
p1
```

### Filter chromosomes with some degree of duplication
To create a version of this notebook with and without the filter
This version is not filtered
```{r, echo = FALSE, eval = FALSE}
#| code-fold: false
coverage_threshold <- 20
depth_threshold <- 1.4
```

```{r, echo = FALSE, eval = FALSE}
chrom_metrics <- filter(chrom_metrics, 
    coverage_percent_duplication >= coverage_threshold | 
    norm_chrom_median >= depth_threshold)
```

### Ploidy and Type (Partial / Full)
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_point(alpha = 0.1)+
        geom_point(aes(color = ploidy, shape = type))+
        scale_color_brewer(name = "Ploidy", palette = "Set1") +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom", legend.direction = "horizontal")+
        labs(y = "Normalized Chromosome Median",
            x = "Percent of Chromosome Covered by Duplications",
            shape = "Type")
```

### Percent covered by deletion
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = norm_chrom_median))+
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_point(aes(color = coverage_percent_deletion, shape = type))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
<details>
<summary> Extra </summary> 
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = coverage_percent_deletion))+
        geom_point(aes(color = norm_chrom_median))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(y = norm_chrom_median, x = coverage_percent_deletion))+
        geom_point(aes(color = coverage_percent_duplication))+
        scale_color_viridis_c(direction = -1) +
        theme_minimal() +
        theme(legend.position = "bottom")
```
</details>

### Number of detected duplicated CNV regions per 100kb
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = norm_chrom_median))+
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_point(aes(color = n_regions_100kb_duplication,shape = type))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
<details>
<summary> Extra </summary> 
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = n_regions_100kb_duplication))+
        geom_point(aes(color = norm_chrom_median))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(y = norm_chrom_median, x = n_regions_100kb_duplication))+
        geom_point(aes(color = coverage_percent_duplication))+
        scale_color_viridis_c(direction = -1) +
        theme_minimal() +
        theme(legend.position = "bottom")
```
</details>

### Span
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_point(aes(color = span_percent_duplication,shape = type)) +
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
<details>
<summary> Extra </summary> 
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = span_percent_duplication))+
        geom_point(aes(color = norm_chrom_median))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(y = norm_chrom_median, x = span_percent_duplication))+
        geom_point(aes(color = coverage_percent_duplication))+
        scale_color_viridis_c(direction = -1) +
        theme_minimal() +
        theme(legend.position = "bottom")
```
</details>

### Difference between Span and Coverage
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_point(aes(color = diff_span_cov_duplication,shape = type)) +
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
<details>
<summary> Extra </summary> 
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = diff_span_cov_duplication))+
        geom_point(aes(color = norm_chrom_median))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(y = norm_chrom_median, x = diff_span_cov_duplication))+
        geom_point(aes(color = coverage_percent_duplication))+
        scale_color_viridis_c(direction = -1) +
        theme_minimal() +
        theme(legend.position = "bottom")
```
</details>

### Median depth of the windows that are part of the duplicated CNV regions
```{r}
max_color_range <- paste("2.7 -", max(chrom_metrics$norm_depth_median_duplication, na.rm = TRUE))
colors <- c("#FDE725FF","#95D840FF","#29AF7FFF","#287D8EFF","#440154FF" )
names(colors) <-c("0 - 1.5", "1.6 - 2", "2.1 - 2.3", "2.4 - 2.6", max_color_range)
ggplot(filter(chrom_metrics, smooth_depth_median_duplication < 4), 
            aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_point(aes(color = cut(norm_depth_median_duplication, 
                            breaks = c(-Inf, 1.5, 2, 2.3, 2.6, Inf), 
                            labels = names(colors)),
                            shape = type),
                            alpha = 1) +
        scale_color_manual(values = colors, 
                            name = "Median Depth of dupCNV\nWindows in Chromosome")+
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom", legend.direction = "horizontal")+
        labs(y = "Normalized Chromosome Median",
            x = "Percent of Chromosome Covered by Duplications")

```
<details>
<summary> Extra </summary> 
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = norm_depth_median_duplication))+
        geom_point(aes(color = norm_chrom_median))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(y = norm_chrom_median, x = norm_depth_median_duplication))+
        geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "solid")+
        geom_point(aes(color = coverage_percent_duplication))+
        scale_color_viridis_c(direction = -1) +
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = norm_depth_median_duplication))+
        geom_point(aes(color = norm_chrom_median))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        scale_y_continuous(limits = c(0,5))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(y = norm_chrom_median, x = norm_depth_median_duplication))+
        geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "solid")+
        geom_point(aes(color = coverage_percent_duplication))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(limits = c(0,5))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
</details>

### Smoothed median depth of the windows that are part of the duplicated CNV regions
```{r}
colors <- c("#FDE725FF","#95D840FF","#29AF7FFF","#287D8EFF","#440154FF" )
names(colors) <-c("0 - 1.5", "1.6 - 2", "2.1 - 2.3", "2.4 - 2.6", "2.7 - 4")
ggplot(filter(chrom_metrics, smooth_depth_median_duplication < 4), 
            aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_point(aes(color = cut(smooth_depth_median_duplication, 
                            breaks = c(-Inf, 1.5, 2, 2.3, 2.6, Inf), 
                            labels = names(colors)),
                            shape = type),
                            alpha = 1) +
        scale_color_manual(values = colors, 
                            name = "Smooth Median Depth of dupCNV\nWindows in Chromosome")+
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom", legend.direction = "horizontal")+
        labs(y = "Normalized Chromosome Median",
            x = "Percent of Chromosome Covered by Duplications")

```
<details>
<summary> Extra </summary> 
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = smooth_depth_median_duplication))+
        geom_point(aes(color = norm_chrom_median))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(y = norm_chrom_median, x = smooth_depth_median_duplication))+
        geom_point(aes(color = coverage_percent_duplication))+
        scale_color_viridis_c(direction = -1) +
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = smooth_depth_median_duplication))+
        geom_point(aes(color = norm_chrom_median))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        scale_y_continuous(limits = c(0,5))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(y = norm_chrom_median, x = smooth_depth_median_duplication))+
        geom_point(aes(color = coverage_percent_duplication))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(limits = c(0,5))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
</details>

### Percent of chromosome covered by the largest single-copy region
```{r, message = FALSE, warning = FALSE}
colors <- c("#FDE725FF","#95D840FF","#29AF7FFF","#287D8EFF","#440154FF" )
names(colors) <-c("0 - 5", "6 - 10", "11 -15", "16 - 20", "21 - 100")

ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        # geom_point(aes(color = percent_largest_region_single_copy))+
        # scale_color_viridis_c(direction = -1)+
        geom_point(aes(color = cut(percent_largest_region_single_copy, 
                           breaks = c(0,5,10,15,20,100), 
                            labels = names(colors)),
                            shape = type)) + 
        scale_color_manual(values = colors, 
                            name = "Largest Single-Copy Region")+
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom", legend.direction = "vertical")

```

<details>
<summary> Extra </summary> 
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(x = coverage_percent_duplication, y = percent_largest_region_single_copy))+
        geom_point(aes(color = norm_chrom_median))+
        scale_color_viridis_c(direction = -1) +
        scale_x_continuous(breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom")
```
```{r, message = FALSE, warning = FALSE}
ggplot(chrom_metrics, aes(y = norm_chrom_median, x = percent_largest_region_single_copy))+
        geom_point(aes(color = coverage_percent_duplication))+
        scale_color_viridis_c(direction = -1) +
        theme_minimal() +
        theme(legend.position = "bottom")
```
</details>

### Names  
```{r, message = FALSE, warning = FALSE}
ggplot(filter(chrom_metrics, coverage_percent_duplication >= 20 | norm_chrom_median >= 1.4), 
                aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_point(alpha = 0.5, aes(color = lineage,shape = type))+
        geom_text_repel(aes(label = sample), size = 3, max.overlaps = 10) +
        scale_x_continuous(limits = c(0,100), breaks = seq(0,100, by = 10))+
        theme_minimal() +
        theme(legend.position = "bottom", legend.direction = "horizontal")+
        labs(y = "Normalized Chromosome Median",
            x = "Percent of Chromosome Covered by Duplications")
```
Names, lower left
```{r, message = FALSE, warning = FALSE}
ggplot(filter(chrom_metrics, coverage_percent_duplication >= 20 | norm_chrom_median >= 1.4), 
                aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_point(alpha = 0.5, aes(color = lineage,shape = type))+
        geom_text_repel(aes(label = paste(sample, chromosome)), size = 3, max.overlaps = 10) +
        scale_x_continuous(limits = c(0,50),breaks = seq(0,50, by = 5))+
        scale_y_continuous(limits = c(1, 1.6))+
        theme_minimal() +
        theme(legend.position = "bottom", legend.direction = "horizontal")+
        labs(y = "Normalized Chromosome Median",
            x = "Percent of Chromosome Covered by Duplications")
```
Names, middle right
```{r, message = FALSE, warning = FALSE}
ggplot(filter(chrom_metrics, coverage_percent_duplication >= 20 | norm_chrom_median >= 1.4), 
                aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_point(alpha = 0.5, aes(color = lineage,shape = type))+
        geom_text_repel(aes(label = paste(sample, chromosome)), size = 3, max.overlaps = 10) +
        scale_x_continuous(limits = c(50,100),breaks = seq(50,100, by = 5))+
        scale_y_continuous(limits = c(1.4,2))+
        theme_minimal() +
        theme(legend.position = "bottom", legend.direction = "horizontal")+
        labs(y = "Normalized Chromosome Median",
            x = "Percent of Chromosome Covered by Duplications")
```
Names, upper right
```{r, message = FALSE, warning = FALSE}
ggplot(filter(chrom_metrics, coverage_percent_duplication >= 20 | norm_chrom_median >= 1.4), 
                aes(x = coverage_percent_duplication, y = norm_chrom_median)) +
        geom_hline(yintercept = c(1, 2), color = "black", linetype = "solid") +
        geom_line(data = trend, aes(x= p_dup, y = median_depth), color = "gray")+ # Line of simulated data
        geom_point(alpha = 0.5, aes(color = lineage,shape = type))+
        geom_text_repel(aes(label = paste(sample, chromosome)), size = 3, max.overlaps = 20) +
        scale_x_continuous(limits = c(50,100),breaks = seq(50,100, by = 5))+
        scale_y_continuous(limits = c(2,2.6))+
        theme_minimal() +
        theme(legend.position = "bottom", legend.direction = "horizontal")+
        labs(y = "Normalized Chromosome Median",
            x = "Percent of Chromosome Covered by Duplications")
```


## Filter out partial duplications
```{r}
#| code-fold: false
coverage_threshold <- 80
```
```{r}
duplications_full <- filter(chrom_metrics, coverage_percent_duplication >= coverage_threshold)
```

## Summary tables

<p style="text-align:center;">**Number of samples with duplications in each lineage**</p>

```{r, message=FALSE}
dup_lineage <- duplications_full %>%
    group_by(lineage) %>%
    summarise(aneuploid = n_distinct(sample), samples_in_lineage = first(samples_in_lineage))%>%
    mutate(euploid = samples_in_lineage - aneuploid)%>%
    pivot_longer(cols = c(aneuploid, euploid), names_to = "type", values_to = "n_samples")%>%
    mutate(percent = round((n_samples / samples_in_lineage) * 100, 1))%>%
    mutate(dup_samples = ifelse(type == "aneuploid", n_samples, ""))
dup_lineage$type <- factor(dup_lineage$type, levels = c("euploid","aneuploid"))
                
dup_lineage %>%
        filter(type == "aneuploid")%>%
        select(lineage, n_samples, percent, samples_in_lineage)
```

<p style="text-align:center;">**Number of samples with duplications in each lineage-source**</p>

```{r, message=FALSE}

metadata_source_lineage <- metadata %>%
        group_by(lineage,source,samples_in_lineage) %>%
        summarize(n_lin_source = n_distinct(sample))


dup_lineage_source <- duplications_full %>%
        group_by(lineage,source) %>%
        summarise(aneuploid = n_distinct(sample))%>%
        right_join(metadata_source_lineage, by = c("lineage", "source"))%>%
        replace_na(list(aneuploid = 0))%>%
        arrange(lineage, source)%>%
        mutate(euploid = n_lin_source - aneuploid)%>%
        select(lineage, source, aneuploid, euploid, n_lin_source, samples_in_lineage)%>%
        pivot_longer(cols = c(aneuploid, euploid), names_to = "type", values_to = "n_samples")%>%
        mutate(percent_samples = round((n_samples / samples_in_lineage) * 100, 1),
                percent_source = round((n_lin_source / samples_in_lineage)*100,1),
                dup_samples = ifelse(type == "aneuploid" & source == "Clinical", n_samples, ""),
                type = str_to_title(type),
                category = paste(source, type, sep = " "))

dup_lineage_source$type <- factor(dup_lineage_source$type, levels = c("Euploid","Aneuploid"))
dup_lineage_source$source <- factor(dup_lineage_source$source, levels = c("Environmental","Clinical"))
dup_lineage_source$category <- factor(dup_lineage_source$category, levels =c("Environmental Euploid", "Environmental Aneuploid","Clinical Euploid","Clinical Aneuploid"))

source <- dup_lineage_source%>%
        select(lineage, source, n_lin_source, percent_source)%>%
        distinct
                
dup_lineage_source  %>%
        filter(type == "Aneuploid")%>%
        select(lineage, source, n_samples, percent_samples, n_lin_source)
```


<p style="text-align:center;">**Number of duplicated chromosomes per sample**</p>

```{r,message=FALSE}
dup_sample <- duplications_full %>%
    group_by(dataset,lineage, sample, strain, source) %>%
    summarise(n_chroms = n_distinct(chromosome), 
            chromosomes = paste(chromosome, collapse = ", ")) %>%
    arrange(desc(n_chroms))
dup_sample
```
  

<p style="text-align:center;">**Number of samples with duplications in each group of dataset-lineage-chromosome**</p>

```{r, message=FALSE}
dup_dataset_lineage_chromosome <- duplications_full %>%
    group_by(dataset,lineage, chromosome) %>%
    summarise(n_samples = n_distinct(sample), 
        samples_in_dataset_lineage = first(samples_in_dataset_lineage))%>%
    mutate(percent_samples = round((n_samples / samples_in_dataset_lineage) * 100, 1))%>%
    select(dataset,lineage, chromosome, n_samples, samples_in_dataset_lineage, percent_samples)%>%
    arrange(chromosome, desc(lineage), desc(n_samples))
dup_dataset_lineage_chromosome
```
  

<p style="text-align:center;">**Number of samples with duplications in each group of lineage-chromosome**</p>

```{r,message=FALSE}
dup_lineage_chromosome <- duplications_full%>%
    group_by(lineage, chromosome) %>%
    summarise(n_samples = n_distinct(sample), 
        samples_in_lineage = first(samples_in_lineage))%>%
    mutate(percent_samples = round((n_samples / samples_in_lineage) * 100, 1))%>%
    select(lineage, chromosome, n_samples, samples_in_lineage,percent_samples)%>%
    arrange(chromosome, desc(lineage), desc(n_samples))
dup_lineage_chromosome 
```
 

<p style="text-align:center;">**Number of samples with duplications in each group of lineage-dataset**</p>

```{r, message=FALSE}
dup_lineage_dataset <- duplications_full%>%
    group_by(dataset,lineage) %>%
    summarise(n_samples = n_distinct(sample), samples_in_dataset_lineage = first(samples_in_dataset_lineage))%>%
    mutate(percent_samples = round((n_samples / samples_in_dataset_lineage) * 100, 1))%>%
    select(lineage, n_samples, samples_in_dataset_lineage, percent_samples)%>%
    arrange(desc(lineage), desc(n_samples))
dup_lineage_dataset

```
  

<p style="text-align:center;">**Number of samples with duplications in each chromosome**</p>

```{r, message=FALSE}
dup_chromosome <- duplications_full %>%
    group_by(chromosome) %>%
    summarise(n_samples = n_distinct(sample), total_samples = first(total_samples))%>%
    mutate(percent_samples = round((n_samples / total_samples) * 100, 1))%>%
    select(chromosome, n_samples,total_samples, percent_samples)%>%
    arrange(chromosome, desc(n_samples))
dup_chromosome 
```
