---
title: "Detect chromosomal duplications"
author: "Claudia Zirión-Martínez"
date: "2025-02-13"
---
## Setup

```{r, echo = FALSE, eval = FALSE}
setwd("/FastData/czirion/Crypto_Diversity_Pipeline/analyses")
```
```{r, output=FALSE}
library(tidyverse)
library(ggbeeswarm)
library(ggtree)
library(ggtreeExtra)
library(ape)
library(phytools)
library(ggnewscale)
library(RColorBrewer)
```
Paths
```{r}
chrom_length_path <- "data/processed/chromosome_lengths.tsv"
metadata_path <- "data/processed/metadata_ashton_desj_all_weavepop_H99.csv"

chrom_names_path <- "../Crypto_Desjardins_Ashton/results_clean/02.Dataset/chromosomes.csv"
cnv_calls_path <- "../Crypto_Desjardins_Ashton/results_clean/02.Dataset/cnv/cnv_calls.tsv"
repeats_path_prefix <- "../Crypto_Desjardins/results_clean/03.References/"
depth_by_chrom_good_desj_path <- "../Crypto_Desjardins/results_clean/04.Intermediate_files/02.Dataset/depth_quality/depth_by_chrom_good.tsv"
depth_by_chrom_good_ashton_path <- "../Crypto_Ashton/results_clean/04.Intermediate_files/02.Dataset/depth_quality/depth_by_chrom_good.tsv"

duplications_path <- "results/tables/duplications_putative.tsv"
```
## Metadata
Use the metadata table that has all the samples included in the final Crypto_Desjardins_Ashton dataset (n = 1055) and H99 .  
```{r, message=FALSE, warning=FALSE}
metadata <- read.delim(
    metadata_path,
    header=TRUE,
    sep=",",
    stringsAsFactor = TRUE)
```
Select needed columns, remove H99 and get the number of samples per dataset and lineage, per lineage, and total.
```{r}
metadata <- metadata %>%
    select(sample, strain, source, lineage, dataset, vni_subdivision)%>%
    filter(!strain == "H99") %>%
    group_by(dataset, lineage)%>%
    mutate(samples_in_dataset_lineage = n_distinct(sample))%>%
    ungroup() %>%
    group_by(lineage)%>%
    mutate(samples_in_lineage = n_distinct(sample))%>%
    ungroup()%>%
    mutate(total_samples = n_distinct(sample))
```

## Chromosome names and length

Get the nice chromosome names 
```{r}
chromosome_names = read.delim(
    chrom_names_path,
    header=TRUE, sep=",")
chromosome_names <- chromosome_names %>%
    mutate(chromosome = str_pad(chromosome, 2, pad = "0"))%>%
    mutate(chromosome = as.factor(chromosome))
levels(chromosome_names$chromosome) <- paste("chr", chromosome_names$chromosome, sep="")
```
Get the chromosome lengths

```{bash, eval=FALSE}
# /usr/bin/bash
tail -n +2 ../Crypto_Desjardins/config/chromosomes.csv | \
    cut -d',' -f1 | sort | uniq | while read line \
do
seqkit fx2tab ../Crypto_Desjardins/data/references/$line.fasta \
    -l -i -n >> data/processed/chromosome_lengths.tsv
done

```
```{r}
chromosome_lengths = read.delim(
    chrom_length_path,
    header=FALSE, 
    col.names=c("accession", "length"), 
    sep="\t")
```

## Get metrics of chromosomes from called CNVs

```{r}
cnv_calls <- read.delim(
    cnv_calls_path,
    header=TRUE, sep="\t")
```

Define a threshold of allowed fraction of repetitive sequences in called CNV regions.
```{r}
repeat_fraction_threshold <- 0.2
```

Get the percentage of the size of each chromosome that is part of called duplications `percent_cnv_size` and
the percentage of the size of the chromosome that is in between the leftmost and rightmost duplicated regions `percent_size_covered`.
```{r, message=FALSE}
cnv_percent <- cnv_calls %>%
    filter(repeat_fraction < repeat_fraction_threshold) %>%
    filter(cnv == "duplication")%>%
    group_by(accession, sample, cnv) %>%
    summarise(total_cnv_size = sum(region_size),
                n_cnvs = n(),
                first = min(start),
                last = max(end),
                mean_cnv_depth = round(mean(smooth_depth),2)) %>%
    left_join(chromosome_lengths, by="accession") %>%
    left_join(chromosome_names, by="accession") %>%
    left_join(metadata, by=c("sample", "lineage")) %>%
    mutate(percent_cnv_size = round((total_cnv_size / length) * 100, 2),
            size_covered = last - first,
            percent_size_covered = round((size_covered / length) * 100, 2))%>%
    mutate(chromosome = as.factor(chromosome)) %>%
    select(dataset,lineage, samples_in_lineage, samples_in_dataset_lineage,
    total_samples, sample,strain, source, accession, chromosome, 
    percent_cnv_size, percent_size_covered, mean_cnv_depth, n_cnvs)
```
```{r, echo=FALSE}
#| fig-height: 7
#| fig-width: 9
#| fig-align: center
ggplot(cnv_percent)+
    geom_quasirandom(aes(x=chromosome, y=percent_cnv_size, color = source))+
    facet_grid(lineage~chromosome, scales="free")+
    labs(x="Chromosome", y="Percent of chromosome duplicated (%)", title=paste("Distribution of the size of duplications with ", repeat_fraction_threshold, "fraction of repeats allowed"))+
    ylim(0, 100)+
    theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color = "lightgray"),
        panel.grid.minor.y = element_blank(),
        panel.background = element_rect(fill = "white"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(colour = "lightgray", fill=NA, linewidth = 2),
        legend.position = "bottom")
```

Get the percentage of each chromosome covered by repeats to know how much of a chromosome might not have reliable CNV calls.

```{r}
lineages <- unique(metadata$lineage)
repeats_all <- data.frame()
for(lineage in lineages){
    repeats_path <-paste(
        repeats_path_prefix,
        lineage, "/", lineage, "_repeats.bed",
        sep ="")
    repeats <- read.delim(repeats_path, 
        header=FALSE, 
        col.names=c("accession", "start", "end", "repeat_type"), 
        sep="\t")
    repeats$lineage <- lineage
    repeats_all <- rbind(repeats_all, repeats)
}
```
```{r, message=FALSE}
repeats_percent <- repeats_all %>%
    mutate(repeat_size_each = end - start)%>%
    group_by(accession, lineage) %>%
    summarise(repeat_size = sum(repeat_size_each)) %>%
    left_join(chromosome_lengths, by="accession") %>%
    left_join(chromosome_names, by=c("accession","lineage")) %>%
    mutate(percent_repeat_size = round((repeat_size / length) * 100, 2))%>%
    mutate(chromosome = as.factor(chromosome))%>%
    select(lineage, accession, chromosome, percent_repeat_size)
```

```{r, echo=FALSE}
#| fig-height: 5
#| fig-width: 9
#| fig-align: center
ggplot(repeats_percent)+
    geom_col(aes(x=lineage, y=percent_repeat_size, fill = lineage))+
    facet_grid(~chromosome, scales="free")+
    labs(title = "Percentage of each chromosome covered by repeats", x="Lineage", y="Percent of chromosome covered by repeats (%)")+
    theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color = "lightgray"),
        panel.grid.minor.y = element_blank(),
        panel.background = element_rect(fill = "white"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(colour = "lightgray", fill=NA, linewidth = 2),
        legend.position = "none")
```
Most chromosomes have repeats in between 5 and 10% of the size. In VNBII it is closer to 15% for some chormosomes.

## Get the chromosome median depth as another measure to dectect duplicated chromosomes

```{r}
depth_by_chrom_good_desjardins <- read.delim(
    depth_by_chrom_good_desj_path,
    header=TRUE, sep="\t")
depth_by_chrom_good_ashton <- read.delim(
    depth_by_chrom_good_ashton_path,
     header=TRUE, sep="\t")
depth_by_chrom_good <- rbind(depth_by_chrom_good_desjardins, depth_by_chrom_good_ashton)
depth_by_chrom <- depth_by_chrom_good %>%
    select(sample, accession, norm_chrom_median)
```

## Join CNV metrics with normalized chromosome depth

```{r}
cnv_and_depth <- left_join(depth_by_chrom, cnv_percent, by = c("sample", "accession" ))
```
Make a threshold for the size of the chromosome that must be in called duplicated regions and a threshold for the normalized chromosome median depth.
```{r}
percent_size_threshold <- 80
depth_threshold <- 1.55
```

Keep, as putative duplications, the chromosomes whose metrics are above any of the thresholds.
```{r}
duplications <- cnv_and_depth %>%
    filter(norm_chrom_median > depth_threshold | percent_cnv_size > percent_size_threshold)%>%
    select(dataset, lineage, samples_in_lineage, samples_in_dataset_lineage, 
        total_samples, source, strain, sample, chromosome, norm_chrom_median,
        mean_cnv_depth, percent_cnv_size, percent_size_covered, n_cnvs)
```
```{r}
head(select(duplications, strain, chromosome, 
    norm_chrom_median, mean_cnv_depth, percent_cnv_size, 
    percent_size_covered, n_cnvs))
```

Write the duplications table to a table to use it in the scripts `duplications_gather_plots.xsh`
and `duplications_polish.qmd` to polish the results. 
```{r, output=FALSE}
write_tsv(duplications, duplications_path)
```
Plot all metrics of putative duplications.

```{r}
duplications_long <- duplications  %>%
    arrange(percent_cnv_size) %>%
    mutate(chrom_sample = paste(chromosome, sample, sep="_"))
duplications_long$chrom_sample <- factor(duplications_long$chrom_sample, 
    levels = duplications_long$chrom_sample)
duplications_long <- pivot_longer(duplications_long, 
    cols = c("norm_chrom_median", "mean_cnv_depth",
        "percent_cnv_size", "percent_size_covered", 
        "n_cnvs"), 
    names_to = "variable", 
    values_to = "value")
duplications_long$variable <- factor(duplications_long$variable,
    levels = c("norm_chrom_median", "mean_cnv_depth",
        "percent_cnv_size", "percent_size_covered", 
        "n_cnvs"))
```
```{r, echo = FALSE}
#| fig-height: 9
#| fig-width: 9
#| fig-align: center
ggplot(duplications_long, aes(x = chrom_sample, y = value))+
        geom_point()+
        facet_grid(scales="free", rows= vars(variable))+
        theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")
```

## Get multiple summary tables

Number of duplicated chromosomes per sample.
```{r,message=FALSE}
dup_sample <- duplications %>%
    group_by(dataset,lineage, sample, strain, source) %>%
    summarise(n_chroms = n_distinct(chromosome), 
            chromosomes = paste(chromosome, collapse = ", ")) %>%
    arrange(desc(n_chroms))
dup_sample
```

Number of samples with duplications in each group of dataset-lineage-chromosome.
```{r, message=FALSE}
dup_dataset_lineage_chromosome <- duplications %>%
    group_by(dataset,lineage, chromosome) %>%
    summarise(n_samples = n_distinct(sample), 
            samples_in_dataset_lineage = first(samples_in_dataset_lineage))%>%
    mutate(percent_samples = round((n_samples / samples_in_dataset_lineage) * 100, 1))%>%
    select(dataset,lineage, chromosome, 
            n_samples, samples_in_dataset_lineage, percent_samples)%>%
    arrange(chromosome, desc(lineage), desc(n_samples))
dup_dataset_lineage_chromosome
```

Number of samples with duplications in each group of lineage-chromosome.
```{r,message=FALSE}
dup_lineage_chromosome <- duplications%>%
    group_by(lineage, chromosome) %>%
    summarise(n_samples = n_distinct(sample), samples_in_lineage = first(samples_in_lineage))%>%
    mutate(percent_samples = round((n_samples / samples_in_lineage) * 100, 1))%>%
    select(lineage, chromosome, n_samples, samples_in_lineage,percent_samples)%>%
    arrange(chromosome, desc(lineage), desc(n_samples))
dup_lineage_chromosome 
```

Number of samples with duplications in each group of lineage-datset.
```{r, message=FALSE}
dup_lineage_dataset <- duplications%>%
    group_by(dataset,lineage) %>%
    summarise(n_samples = n_distinct(sample), 
        samples_in_dataset_lineage = first(samples_in_dataset_lineage))%>%
    mutate(percent_samples = round((n_samples / samples_in_dataset_lineage) * 100, 1))%>%
    select(lineage, n_samples, samples_in_dataset_lineage, percent_samples)%>%
    arrange(desc(lineage), desc(n_samples))
dup_lineage_dataset

```

Number of samples with duplications in each chromosome.
```{r, message=FALSE}
dup_chromosome <- duplications %>%
    group_by(chromosome) %>%
    summarise(n_samples = n_distinct(sample), total_samples = first(total_samples))%>%
    mutate(percent_samples = round((n_samples / total_samples) * 100, 1))%>%
    select(chromosome, n_samples,total_samples, percent_samples)%>%
    arrange(chromosome, desc(n_samples))
dup_chromosome 
```