---
title: "Ggtree notes"
author: "Claudia Zirión-Martínez"
date: "2025-05-29"
---
## Setup

```{r, echo = FALSE, eval =FALSE}
setwd("/FastData/czirion/WeavePop_Cneoformans/analyses/")
```
Libraries
```{r, output=FALSE}
library(tidyverse)
library(ggtree)
library(ggtreeExtra)
library(ape)
library(ggnewscale)
library(RColorBrewer)
library(svglite)
```

Paths 
```{r}
metadata_path <- 
    "data/processed/metadata_ashton_desj_all_weavepop_H99.csv"
duplications_path <- 
    "results/tables/duplications.tsv"
merged_tree_path <- 
    "data/processed/tree_merged.newick"
```

Include color palettes from R script
```{r}
source("scripts/metadata_colors.R")
```
Example of color palette objects
```{r}
lineage_colors
```

## Tree
Read tree
```{r, message = FALSE, warning = FALSE}
tree_original <- read.tree(merged_tree_path)
```

## Metadata

Read and organize metadata. 

To add metadata to a ggtree object (`ggtree(tree) %<+% metadata`), the **first column** of 
the metadata table must contain the labels in `tree$tip.label`.

```{r, message = FALSE, warning = FALSE}
metadata <- read.csv(
    metadata_path,
    header = TRUE) %>%
    select(strain, sample, source, lineage, country_of_origin, dataset)
head(metadata)
```

## Manipulate tree
Remove tips that are not in `metadata$strain`
```{r, message = FALSE, warning = FALSE}
tree <- drop.tip(tree_original, setdiff(tree_original$tip.label, metadata$strain))
```

### Color tips 

Color the points of the tips with `geom_tippoint()`
This works even when not all tips are in the metadata table, NA values are used when needed.
```{r, message = FALSE, warning = FALSE}
ggtree(tree_original, layout = "circular") %<+%  metadata +
    geom_tippoint(aes(color = lineage), size = 2)+
    scale_color_manual(name="Lineage", values= lineage_colors)
```

### Color labels

Add the tip labels (strain) and color them by a column in the metadata.
```{r, message = FALSE, warning = FALSE}
ggtree(tree, layout = "circular") %<+%  metadata +
    geom_tiplab(aes(color = source), size = 2, align = TRUE)+
    scale_color_manual(name="Source", values= source_colors)
```

### Identify basal nodes of clades

To identify the most basal node of each lineage, get the Most Recent Common Ancestor 
of a pair of strains in the lineage, they have to be two strains whose MRCA is at the base of the 
clade. You can also use this to identify any clade of interest.
```{r, message = FALSE, warning = FALSE}
VNI_node <- getMRCA(tree, c("Tu241-1","UI_31647-2"))
VNII_node <- getMRCA(tree, c("C2","C12"))
VNBI_node <- getMRCA(tree, c("Tu229-1","Ftc267-2"))
VNBII_node <- getMRCA(tree, c("MW-RSA3321","MW-RSA3179"))

nodes_lineages <- data.frame(
    lineage = c("VNI", "VNII", "VNBI", "VNBII"),
    mrca = c(VNI_node, VNII_node, VNBI_node, VNBII_node)
)
```
Get the node number of the MRCA of each VNI sublineage
```{r, message = FALSE, warning = FALSE}
VNIa4_node <- getMRCA(tree, c("04CN-30-008","UI_31647-2"))
VNIa5_node <- getMRCA(tree, c("BMD852","04CN-64-014"))
VNIa93_node <- getMRCA(tree, c("04CN-65-080","04CN-65-002"))
VNIa32_node <- getMRCA(tree, c("BMD942","BMD2801"))
VNIaX_node <- getMRCA(tree, c("Bt48","04CN-63-007"))
VNIaY_node <- getMRCA(tree, c("Bt6","Bt138"))
VNIb_node <- getMRCA(tree, c("04CN-65-096","MW-RSA722"))
VNIc_node <- getMRCA(tree, c("Bt20","Bt11"))

nodes_sublineages <- data.frame(
    sublineage = c(
                "VNIa-4", "VNIa-5", "VNIa-93",
                "VNIa-32", "VNIa-X", "VNIa-Y",
                "VNIb", "VNIc"),
    mrca = c(
            VNIa4_node, VNIa5_node, VNIa93_node,
            VNIa32_node, VNIaX_node, VNIaY_node,
            VNIb_node, VNIc_node))
```

To print the tree with all the node labels:
```{r, message = FALSE, warning = FALSE}
ggtree(tree, layout = "circular")+
    geom_text(aes(label = node), size = 3, vjust = -0.5)
```

### Add clade labels
```{r, message = FALSE, warning = FALSE}
ggtree(tree, layout = "circular") +
    geom_cladelab(data = nodes_lineages, 
                mapping = aes(node = mrca, label = lineage, color = lineage))+
    scale_color_manual(name="Lineage", values= lineage_colors)

```

## Add node labels
```{r, message = FALSE, warning = FALSE}
ggtree(tree, layout = "circular") +
        geom_text(aes(label = nodes_lineages$lineage[match(node, nodes_lineages$mrca)]), 
                        size = 4, , fontface = "bold",
                        hjust = 1.25, vjust = -0.5)
```

## Color branches

To color the branches of clades, create groups in the tree object with `groupClade()`.
```{r, message = FALSE, warning = FALSE}
tree <- groupClade(tree, setNames(nodes_sublineages$mrca, nodes_sublineages$sublineage))
```
And color the tree by `aes(color = group)`. 
Use `scale_color_manual()` with a palette to specify the colors of 
the clades and avoid coloring the branches that are not part 
of the selected clades. 
```{r, message = FALSE, warning = FALSE}
ggtree(tree, aes(color = group), layout = "circular", 
    size = 1) +
  scale_color_manual(
    name = "Sublineage",
    values = sublineage_colors
  )
```

## Color branches and tips and use node labels
```{r, message = FALSE, warning = FALSE}
ggtree(tree, aes(color = group), 
        layout = "circular", size = 1) %<+%  metadata +
    scale_color_manual(
        name = "Sublineage",
        values = sublineage_colors)+
    new_scale_color()+
    geom_text(aes(label = nodes_lineages$lineage[match(node, nodes_lineages$mrca)]), 
                        size = 4, , fontface = "bold",
                        hjust = 1.25, vjust = -0.5)+
    geom_tippoint(aes(color = source), size = 1)+
    scale_color_manual(name = "Source", values = source_colors)
```

```{r, message = FALSE, warning = FALSE}
ggtree(tree, layout = "circular") %<+%  metadata +
    geom_cladelab(data = nodes_sublineages, 
                mapping = aes(node = mrca, label = sublineage, color = sublineage),
                align = TRUE, face = "bold")+
    scale_color_manual(name="Sublineage", values= sublineage_colors)+
    geom_text(aes(label = nodes_lineages$lineage[match(node, nodes_lineages$mrca)]), 
                        size = 6, , fontface = "bold",
                        hjust = 1.25, vjust = -0.5)+
    new_scale_color() +
    geom_tippoint(aes(color = source), size = 2)+
    scale_color_manual(name = "Source", values = source_colors)
```

## Add data as heatmap

Create a dataframe where the rownames are the tree's tip labels, each column is 
a column of the heatmap and the values are the colors in the heatmap.
If using categorical values for the heatmap and their order in the legend is important,
convert the column to factor and set the order of the levels.
```{r, message = FALSE, warning = FALSE}
country <- metadata %>%
    select(strain, country_of_origin)%>%
    column_to_rownames("strain")

country$country_of_origin <- factor(country$country_of_origin,
                                levels = names(country_colors))
```

To add a heamap, the function `gheatmap()` takes an object made with `ggtree()` and
add the dataframe with the matrix to use in the heatmap.
Use the `offset` option to separate the heatmap from the tree.
Use the `guides()` function to order the legends.
Use the option `limits` in `scale_fill_manual` to set the order of the legend labels 
to match the order of the levels in the factor. 
```{r, message = FALSE, warning = FALSE}
#| fig-height: 9
#| fig-width: 7
#| fig-align: center
p <- ggtree(tree, branch.length = "none")  %<+%  metadata
gheatmap(p, country, width=0.1, colnames = FALSE,  
            colnames_position = "top", 
            offset=7,
            font.size = 6) +
    geom_tiplab(aes(color = lineage), size = 3)+
    scale_color_manual(name = "Lineage", values = lineage_colors)+
    scale_fill_manual(values=country_colors, 
                      name="Country", na.translate=FALSE,
                      limits = levels(country$country_of_origin))+
    guides(color = guide_legend(order = 1), fill = guide_legend(order = 2))+
    theme(legend.position = "right")
```

